---
title: "STAT 547 Class Meeting 02 Worksheet"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

```{r}
suppressPackageStartupMessages(library(tidyverse))
library(gapminder)
library(testthat)
```

## Resources

Today's lesson has been drawn from the following resources:

1. Mostly [stat545.com: character data](http://stat545.com/block028_character-data.html)
    - See the ["Resources" section](http://stat545.com/block028_character-data.html#resources) for a more comprehensive listing of resources based on the character problem you're facing. 
2. [Older stat545 notes](http://stat545.com/block022_regular-expression.html)
3. [r4ds: strings](https://r4ds.had.co.nz/strings.html).
4. [`stringr` vignette](https://cran.r-project.org/web/packages/stringr/vignettes/stringr.html)

## Basic String Manipulation

__Goal__: Go over some basic functionality of `stringr`. 

There's that famous sentence about the quick brown fox that contains all letters of the alphabet, although I don't quite remember the sentence. Demo: Check to see if it's in the `sentences` data. Try:

`str_detect(string, pattern)` #search the string for a particular pattern
`str_subset(string, pattern)`

```{r}
(fox <- str_subset(sentences, pattern = "fox")) #argument one is what we are looking in. The second argument is what we are looking for.
```

Not quite the sentence I was thinking of. How many words does it contain? Use `str_split(string, pattern)`, noting its output (list).

```{r}
str_split(fox, pattern = " ")[[1]] %>% 
  length()#this will split of the fox string by spaces. Each word is now it's own thing, and you get a list. Then the [[1]] and legnth() will count it for you.

str_split(sentences, pattern = " ") %>% 
  head()
```

[[1]] -> give me the contents of the first car in a train
[1] -> give me the first car of a train
`[[`(1) -> this also gives you the contents of the first car

Exercise: does this sentence contain all letters of the alphabet? Hints: 

- Split by `""`. 
- Consider putting all in lowercase with `str_to_lower()`.
- Use the base R `table()` function.

```{r}
fox %>% 
  str_split("") %>% # we are splitting everything up by each letter and making each letter a character in itself. Run this code to see what happens.
  `[[`(1) %>% 
  str_to_lower() %>% 
  table()


alist <- list()
alist 

```


Working in a data frame? `tidyr` has its own version of this. Here's an example from Resource 1, with the fruit data:

```{r}
tibble(fruit)
tibble(fruit) %>%
  separate(fruit, into = c("pre", "post"), sep = " ") # Makes a new column for each word.
```

Demo: we can substitute, too. Replace the word "fox" with "giraffe" using `str_replace(string, pattern, replacement)`:

```{r}
fox %>% 
    str_replace(pattern = "fox", replacement = "giraffe")
```

Know the position you want to extract/replace? Try `str_sub()`.

`str_pad()` extends each string to a minimum length:

```{r}
fruit %>% 
  head

fruit %>% 
    str_pad(width=7, side="right", pad="$") %>% #adds dollar signs to make everything 7 letters long 
    head()
```

`str_length()` (Not the same as `length()`!)

```{r}
str_length(fruit) # tells us how many characters there are in each vector
length(fruit) # tells us how many types of fruits there are
  
```

`str_c()` for concatenating strings. Check the docs for an excellent explanation using a matrix.

```{r}
str_c(words[1:4], words[5:8], sep=" & ") # first four words of data, and combine with next four words of data. Joined together with &. 
str_c(words[3:4], words[5:8], sep=" & ") # if you don't have enough if the first argument R will recycle the first two words.
str_c(words[3:4], words[5:8], sep=" & ", collapse=", ") # takes all the characters, and puts them together with commas separating
```

There's a (more limited) `tidyr` version. Straight from Resource 1:

```{r}
fruit_df <- tibble(
  fruit1 = fruit[1:4],
  fruit2 = fruit[5:8]
) # make two different columns. Then put them together with unite() after.
fruit_df %>% 
  unite("flavor_combo", fruit1, fruit2, sep = " & ") # brings the two columns together.
```


## Exercise: Populate your Participation Repo

So, you don't want to manually make 12 folders for your participation repo. I hear you. Let's do that by making a character vector with entries `"cm101"`, `"cm102"`, ..., `"cm112"`.

(If you've already done this, it's still a useful exercise!)

### Make Folders

Let's make those folders!

1. Make a character vector with entries `"01"`, `"02"`, ..., `12` with `str_pad()`.

```{r}
(num <- str_pad(string = 1:12, width = 2, side = "left", pad = "0"))

#So we just made a list of characters with numbers in them.

```

2. Use `str_c()` to combine `"cm1"` with the numbers:
    - If your system uses "\" instead of "/", you might need two backslashes.

```{r}
(folders <- str_c("/cm1", num))
```

3. Use `testthat` to check that each entry of `folders` has 5 characters. You might find the base R `all()` function useful.

```{r}
test_that("folder names are length 5.", {
    expect_true(all(str_length(folders) == 6))
})
```

4. BONUS: If applicable, make the folders using `dir.create()`.
    - Note: `dir.create()` requires the full path to be specified. You might find the `here::here()` function useful.
    - This code might work (depending on your directory): `for (folder in folders) dir.create(here::here(folder))`
    - We'll learn how to use `purrr` instead of loops next week.

### Make README's

Now, let's seed the folders with README's. Each of the folders needs a README file in it.

1. Add `/README.md` to the end of the folder names stored in `folders`:

```{r}
(files <- str_c(folders, "/README.md"))
```

2. Make a vector of contents to put in each README. Put a title and body. 
    - Hint: Use `\n` to indicate a new line! This works in graphs, too.
    - So now we are making a bunch of text to put in the README file
    
```{r}
contents <- str_c("# Participation\n\n Participation for Class Meeting", 1:12)
cat(contents[1]) # this is what the first one would look like
```

3. BONUS: Write the README's to file using base R's `write(x, file)`:
    - `for (i in 1:length(files)) write(contents[i], files[i])`
    - There's a better alternative to a loop using `purrr`. Next week's topic!
    - This code might not work, depending on your workind directory and system.

## Regular Expressions (aka regex)

Great resource is [r4ds](https://r4ds.had.co.nz/strings.html#matching-patterns-with-regular-expressions)!

Premable:

- Useful for identifying _patterns_, not exact character specifications.
- Hard to read and write!
- We'll focus on finding _matches_ (the hardest part). You can also use regex to manipulate strings -- but we'll delegate that to [r4ds: strings: tools](https://r4ds.had.co.nz/strings.html#tools).

Staying true to Resource 1, let's work with the gapminder countries:

```{r}
library(gapminder)
countries <- levels(gapminder$country)
```

### The "any character"

Find all countries in the gapminder data set with the following pattern: "i", followed by any single character, followed by "a":

```{r}
str_subset(countries, pattern = "i.a") # so here we are looking for countries with i something a. Note that this is case sensitive so we won't see Italy in this.
```

Here, `.` stands for "any single character".

But, where's Italy? Case-sensitive!

Let's use `str_view_all()` to see the matches:

```{r}
str_view_all(countries, pattern = "i.a") # this will highlight where it found that in the country names.
str_view_all(countries, pattern = "i.a", match=TRUE) # this will only show the ones where it finds the matches
```

Exercise: Canada isn't the only country with three interspersed "a"'s. Find the others. Try both `str_view_all()` and `str_subset()`.

```{r}
str_view_all(countries, pattern = "a.a.a", match=TRUE) 
str_subset(countries, pattern = "a.a.a")
```


Let's define a handy function - this makes it so you don't have to write the match = TRUE all the time.

```{r}
str_view_all_match <- function(countries, pattern) {
    str_view_all(countries, pattern, match=TRUE)
}
str_view_all_match(countries, pattern = "i.a")
```


### The escape

What if I wanted to literally search for countries with a period in the name? Escape with `\`, although R requires a double escape.

```{r}
str_view_all_match(countries, pattern = "\\.") # see all countries with . in it
```

Why does R require a double escape? It does one level of escaping before "executing" the regex. 

- See `?Quotes`
- Try searching for "s\. " (without quotes) in this document (don't forget to select "Regex")

### Character Classes

- `[letters]` matches a single character that's either l, e, t, ..., or s.
- `[^letters]`: anything _but_ these letters. 

See more at: https://r4ds.had.co.nz/strings.html#character-classes-and-alternatives

Note that not all special characters "work" within `[]`, but some do, and do not always carry the same meaning (like `^`)! From said resource, they are:

>  `$` `.` `|` `?` `*` `+` `(` `)` `[` `{`. Unfortunately, a few characters have special meaning even inside a character class and must be handled with backslash escapes: `]` `\` `^` and `-`.
Exercise: Find all countries with three non-vowels next to each other.

```{r}
str_view_all_match(countries, pattern = "[^aeiou]") # this is finding where there is no lowercase vowels

str_view_all_match(countries, pattern = "[^aeiou][^aeiou][^aeiou]") # this is finding where there is no lowercase vowels three for three letters in a row.

#Remember that the ^ is like ! -> means 'not' this.
```


### Or

- Use `|` to denote "or". 
- "And" is implied otherwise, and has precedence.
- Use parentheses to indicate precedence.

Beer or bear?

```{r}
c("bear", "beer", "bar") %>% 
    str_view_all_match(pattern = "be(e|a)r") # find all the words with either bear or beer.
```


### Quantifiers/Repetition

The handy ones are:

- `*` for 0 or more
- `+` for 1 or more
- `?` for 0 or 1

See list at https://r4ds.had.co.nz/strings.html#repetition

Find all countries that have any number of o's (but at least 1) following r:

```{r}
str_view_all_match(countries, "ro+") # r followed by o at least once
```


Find all countries that have exactly two e's next two each other:

```{r}
str_view_all_match(countries, "e{2}") # two e's in a row -> could have just wrote ee
```


Exercise: Find all countries that have either "a" or "e", twice in a row (with a changeover allowed, such as "ae" or "ea"):

```{r}
str_view_all_match(countries, pattern="(a|e)(a|e)") # find countries with a or e twice in a row

str_view_all_match(countries, pattern="(a|e){2}") # find countries with a or e twice in a row
```


### Position indicators

- `^` corresponds to the __beginning__ of the line. If we had this in [^] it would be not what comes after.
- `$` corresponds to the __end__ of the line.

Countries that end in "land":

```{r}
str_view_all_match(countries, pattern = "land$") # just want land if it's at the end
str_view_all_match(countries, pattern = "$") # gives us everything because everything has an end
```

Countries that start with "Ca":

```{r}
str_view_all_match(countries, pattern = "^Ca")
```

Countries without a vowel? The word should start with a non-vowel, continue as a non-vowel, and end:

```{r}
str_view_all_match(countries, "^[^aeiouAEIOU]*$") 
```

### Groups

We can refer to parentheses groups **I dont get this**

```{r}
str_view_all(c("abad", "abbd"), pattern="(a)(b)\\1") # Want an a followed by a single b, and I want what the first group was resolved to.
str_view_all(c("abad", "abbd"), pattern="(a)(b)\\2") # Want an a followed by a single b, and I want what the second group was resolved to.
```

Note that the parentheses are first resolved, THEN referred to. NOT re-executed.

```{r}
str_view_all(c("bananas"), "(.)(.)\\1\\2.*\\1\\2")
```

We can refer to them later in the search, too:

```{r}
str_view_all(c("bananas", "Who can? Bananas can."), "(.)(.)\\1\\2.*\\1\\2")
```


## Final Exercises

Convert `words` to pig latin, which involves:

1. Make the first letter the last letter
    - Get the first letter with `str_sub(string, start, end)`.
2. Remove the first letter from `words`.
    - Hint: leave the `end` argument blank.
3. Add "ay" to the end of the word.
    - Use `str_c()`.
    
```{r}
YOUR_CODE_HERE
```

Find all countries that end in "y"

```{r}
str_view_all_match(countries, "YOUR_REGEX_HERE")
```

Find all countries that have the same letter repeated twice (like "Greece", which has "ee").

```{r}
str_view_all_match(countries, "YOUR_REGEX_HERE")
```

Find all countries that end in two vowels. 

```{r}
str_view_all_match(countries, "YOUR_REGEX_HERE")
```

Find all countries that start with two non-vowels. How is this different from finding all countries that end in _at least_ two non-vowels? Hint: Syria.

```{r}
countries %>% 
    str_to_lower() %>% 
    str_view_all_match("YOUR_REGEX_HERE")
```

Find all countries that have either "oo" or "cc" in them.

```{r}
str_view_all_match(countries, "YOUR_REGEX_HERE")
```