---
title: "cm008 Exercises"
output: github_document
---

## Install `nycflights13` package

```{r,eval=FALSE}
#install.packages("nycflights13")
```

```{r}
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(nycflights13))
```

## Types of mutating join

### Let's join tibbles using four mutating functions:  `left_join`, `right_join`, `inner_join` and `full_join`.

Check out the cheatsheet here: https://www.rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf

### create two tibbles named `a` and `b`

```{r}
(a <- tibble(x1 = LETTERS[1:3], x2 = 1:3))
(b <- tibble(x1 = LETTERS[c(1,2,4)], x3 = c("T", "F", "T")))
```

### left_join: Join matching rows from `b` to `a` by matching "x1" variable

The left join refers to the table on the left. It will only keep the rows from that table. Notice how there is no D row.

```{r}
left_join(a, b, by = "x1")
```


### right_join: Join matching rows from `a` to `b` by matching "x1" variable.

But if we use right join, it will use the rows from the table on the right. Notice that there is a D row, but no C row.

```{r}
right_join(a, b, by = "x1")
```



### inner_join: Join data. Retain only rows in both sets `a` to `b` by matching "x1" variable.

This only finds the rows that have the same values of x1 -> A,B. Gets rid of all the rest.

```{r}
inner_join(a, b, by = "x1")
```


### full_join: Join data. Retain all values, all rows of `a` to `b` by matching "x1"

This keeps all of them

```{r}
full_join(a, b, by = "x1")
```


### what happen if we do not specify `by` option?

```{r}
left_join(a, b)
```


### what happen if we join two different variables (e.g., "x1" to "x3") from two tibbles `a` to `b`?

```{r}
left_join(a, b, by = c("x1" = "x3"))
```


### what happen if two columns of `a` and `c` datasets have the identical colnames?

```{r}
# make data frame c and use inner_join()
(c <- tibble(x1 = c(LETTERS[1:2],"x"), x2 = c(1,4,5)))
inner_join(a, c)
```

## In class practice

`nycflights13` dataset has several tibbles e.g., `flights`, `airports`, `planes`, `weather`.

### 1. Explore `nycflights13` dataset

```{r}
#check the tibbles included in `nycflights13` package
class(flights)

colnames(flights)

colnames(airlines)

colnames(weather)
```

### 2. Drop unimportant variables so it's easier to understand the join results. Also take first 1000 rows to run it faster.

```{r}
flights2 <- flights[1:1000,] %>% 
  select(year, tailnum, carrier, time_hour)
dim(flights2)
#1000 rows and 4 columns/variables
```

### 3. Add airline names to `flights2` from `airlines` dataset.

```{r}
# Which join function to use?
colnames(airlines)
colnames(flights2)

left_join(flights2, airlines, by = "carrier")

flights2 %>% 
  left_join(airlines)

```

### 4. Add `weather` information to the `flights2` dataset by matching "year" and "time_hour" variables.

```{r}
flights2 %>% 
left_join(weather, by = c("year", "time_hour"))


```


### 5. Add `weather` information to the `flights2` dataset by matching only "time_hour" variable

```{r}
flights2 %>% 
left_join(weather, by = "time_hour")
```

## Types of filtering join

### Let's filter tibbles using two filtering functions: `semi_join`, `anti_join` 

### example for `semi_join`: All rows in `a` that have a match in `b`

```{r}
semi_join(a,b)
```


### example for `anti_join`: All rows in `a` that do not have a match in `b`

```{r}
anti_join(a,b)
```


### example of joinin by matching two variables (e.g., "x1", "x2") from both datasets `a` and `c`

```{r}
semi_join(a,c)
```


## Types of Set Operations for two datasets

### Let's use three `set` functions: `intersect`, `union` and `setdiff` 

### create two tibbles named `y` and `z`, similar to Data Wrangling Cheatsheet

```{r}
(y <-  tibble(x1 = LETTERS[1:3], x2 = 1:3))
(z <- tibble(x1 = c("B", "C", "D"), x2 = 2:4))
```

### example for `intersect`: Rows that appear in both `y` and `z`

Will combine rows that appear in both y and z. 

```{r}
intersect(y,z)
```


### example for `union`: Rows that appear in either or both `y` and `z`

Rows that appear in either or both y and z

```{r}
union(y,z)
```


### example for `setdiff`: Rows that appear in `y` but not `z`. __Caution:__ `setdiff` for `y` to `z` and `z` to `y` are different.

Rows that appear in y but not z or vice versa.

```{r}
setdiff(y,z)

setdiff(z,y)
```



### what happen if colnames are differentin `y` and `x`? Is there any error message and why?

```{r}
(x <- tibble(x1 = c("B", "C", "D"), x3 = 2:4))

```

## Types of binding datasets

### Let's bind datasets by rows or column using two binding functions:

### example for `bind_rows`: Append `z` to `y` as new rows

You just add the rows in z to the bottom of y. You can only do this when they have the same type of x2 column.

```{r}
bind_rows(y,z)
```


### example for `bind_cols`: Append `z` to `y` as new columns. __Caution__: matches rows by position. Check colnames after binding.

This will just add two more columns from the z dataframe to the end of y.

```{r}
bind_cols(y,z)
```

### what happen if colnames are different between `y` and `x` datasets?

```{r}
bind_cols(y,x)
```


##Practice Exercises
Practice these concepts in the following exercises. It might help you to first identify the type of function you are applying.


### 1. Filter the rows of `flights2` by matching "year" and "time_hour" variables to `weather` dataset. Use both `semi_join()` and `anti_join()`

```{r}
flights2 %>% 
semi_join(weather, by = c("year", "time_hour"))
```

### 2. Can we apply `set` and `binding` funcions between `flights2` and `weather` datasets. Why and why not?

I don't think so because the variables are differnt.

```{r}

```


### 3. Let's create a tibble `p` with "x1" and "x2" coulmns and have duplicated element in "x1" column. Create another tibble `q` with "x1" and "x3" columns. Then apply `left_join` function `p` to `q` and `q` to `p`. 

```{r}
(p <-  tibble(x1 = LETTERS[c(1:3,1)], x2 = 1:4))
(q <-  tibble(x1 = LETTERS[c(1:3,1)], x3 = 1:4))

left_join()
```

